---
output:
  md_document:
    variant: markdown_github
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-"
)
```

# icews

Get the ICEWS event data from the Dataverse repo at https://dataverse.harvard.edu/dataset.xhtml?persistentId=doi:10.7910/DVN/28075. 

The goal is to eventually develop this into a package that can (a) keep a single local copy of the ICEWS event data in sync with the latest data on Dataverse and (b) be re-used between projects without the need to duplicate the 2-3Gb of data each time.  

## Installation

```{r github-install, eval = FALSE}
library("devtools")
install_github("andybega/icews")
```


## Usage

To download the current data files from Dataverse, use `download_icews`. It accepts a path argument for the download destination. Alternatively, set an environment variable for the destination path. The plan is that in the future, the package will work by keeping a persistent local copy of some sort in sync with the version on Dataverse, including any more recent data releases. But for now, it's just a convenient shortcut to avoid manually downloading from Dataverse. 

(BTW, this takes a while.)

```{r example, eval = FALSE}
library("icews")

Sys.setenv("ICEWS_DATA_DIR" = "~/Dropbox/Work/VForecast/Data/icews_data")
Sys.setenv("ICEWS_USE_DB" = TRUE)

# Alternatively, set to .Renviron file
# library("usethis")
# edit_r_environ()

download_icews()
```

The events come in (zipped) tab-seperated files. To load all of these into memory in a big combined data frame with about 16 million rows (~2.5Gb):

```{r, eval = FALSE}
events <- read_icews()
```

Also included is a dictionary of the CAMEO code for event types. 

```{r}
data("cameo_codes")
head(cameo_codes[, 1:5])
```

